### 2016-02-14 main.tex
Finish edits (Chapter 5)

\ctparttext{This part presents the key novel contributions of the thesis. We develop \emph{flat}
  and \emph{structural} coeffect calculi that capture a wide range of context-aware programming
  languages. We discuss their type systems and semantics together with safety properties.}
\setcounter{part}{1}
\part{Coeffect calculi}

>>>>

\ctparttext{This part presents the key novel contributions of the thesis. We develop \emph{flat}
  and \emph{structural} coeffect calculi that capture a wide range of context-aware programming
  languages discussed in the previous section. We present parameterized coeffect type system, 
  categorical model based on indexed comonads. We use the model to give semantics in terms of 
  translation to a target functional language and we present syntactic safety proof for 
  context-aware languages with implicit parameters and dataflow.}
\setcounter{part}{1}
\part{Coeffect calculi}

### 2016-02-14 text/Chapter05.tex
Finish edits (Chapter 5)

\paragraph{Liveness.}
For example, if we wanted to follow the development in the previous section for liveness, we would
extend the target language with two kinds of expressions, \ident{Dead} representing a dead context 
with no value and \ident{Live} representing a context with a value:
%
\begin{equation*}
\begin{array}{rcl}
e &=& \ldots \sep \ident{Dead} \sep \ident{Live}~e \\
\end{array}  
\end{equation*}
%
The typing rules promote the information about whether a value is available into the type-level 
and so a context carrying a live value is marked as $\ctyp{\cclrd{\ident{L}}}{\tau}$ while a
dead context has a type $\ctyp{\cclrd{\ident{D}}}{\tau}$.
%
\begin{equation*}
\tyrule{live}
  { \Gamma \vdash e:\tau }
  { \Gamma \vdash \ident{Live}~e : \ctyp{\cclrd{\ident{L}}}{\tau} }
\qquad
\tyrule{dead}
  { ~ }
  { \Gamma \vdash \ident{Dead} : \ctyp{\cclrd{\ident{D}}}{\tau} }
\end{equation*}
%
The above example can be encoded using generalized algebraic data types (GADTs) in Haskell. We
define types \ident{D} and \ident{L} to capture the type annotations and we define a type
$\ident{C}~r~a$ where the type parameter $r$ represents a coeffect annotation:
%
\begin{equation*}
\begin{array}{l}
\kvd{data}~\ident{L} \\[-0.2em]
\kvd{data}~\ident{D} \\[0.5em]
\kvd{data}~\ident{C}~r~a~\kvd{where} \\[-0.2em]
\quad \ident{Live} \hspace{0.45em}:: a \rightarrow \ident{C}~\ident{L}~a\\[-0.2em]
\quad \ident{Dead} :: \ident{C}~\ident{D}~a
\end{array}
\end{equation*}
%
The operations of the flat coeffect algebra could then be encoded as type-level operations on
the types $\ident{L}$ and $\ident{D}$ so that, for example $\ident{Seq}~a~b$ is $\ident{D}$ only
if both $a$ and $b$ are $\ident{D}$. The types of the \ident{counit} and \ident{cobind} operations
would encode the coeffect annotations as:

\begin{equation*}
\begin{array}{rcl}
\ident{counit} &::& \ident{C}~\ident{L}~a \rightarrow a \\[-0.2em]
\ident{cobind} &::& (\ident{C}~\ident{r}~a \rightarrow b) \rightarrow \ident{C}~(\ident{Seq}~r~s)~a \rightarrow \ident{C}~\ident{s}~b
\end{array}
\end{equation*}
%
Here, the additional type annotation ensures that \ident{counit} can only be called on a context
that contains a value and so calling the operation is not going to fail. Similarly, the type of the
\ident{cobind} operation now guarantees that if a function (used as the first argument) or the result
require a live context, it will be called with a value $\ident{C}~\ident{L}~a$ that is guaranteed
to contain a value.

In our formalization, this is what guarantees that the reduction rules do 
not get stuck. For liveness, those would look as:
%
\begin{equation*}
\begin{array}{rl}
{\footnotesize(\emph{counit})} &
  \ident{counit}_{\cclrd{ \ident{L} }}~(\ident{Just}~v)   \rightarrow v
\\[0.5em]
{\footnotesize(\emph{cobind-1})} &
  \ident{cobind}_{\cclrd{\ident{L}}, \cclrd{\ident{L}}}~f~(\ident{Live}~v) \rightarrow \ident{Live}~(f~v)
\\
{\footnotesize(\emph{cobind-2})} &
  \ident{cobind}_{\cclrd{\ident{D}}, \cclrd{\ident{L}}}~f~(\ident{Live}~v) \rightarrow \ident{Live}~(f~\ident{Dead})
\\
{\footnotesize(\emph{cobind-3})} &
  \ident{cobind}_{\cclrd{\ident{L}}, \cclrd{\ident{D}}}~f~v \rightarrow \ident{Dead}
\\
{\footnotesize(\emph{cobind-4})} &
  \ident{cobind}_{\cclrd{\ident{D}}, \cclrd{\ident{D}}}~f~v \rightarrow \ident{Dead}
\end{array}
\end{equation*}
%
If we encoded the rules as functions in a dependently-typed language or in Haskell, the type
preservation property would be guaranteed by the type system of the target language. The 
equivalent of the progress property would then be guaranteed by the fact that the the operations
are always defined.

@@

\section{Related and futrue work}
\label{sec:semantics-related}

Most of the related work leading to coeffects has already been discussed in Chapter~\ref{ch:pathways} 
and we covered work related to individual concepts throughout the chapter. In this section, we
do not repeat the discussion present elsewhere. Instead, we discuss one specific question that often 
arises when discussing coeffects and that is \emph{when is a coeffect (not) an effect?}

We start with a quick overview of the ways in which effects and coeffects differ and then
we briefly look at one (but illustrative) example where the two concepts overlap. We focus 
mainly on the equivalence between the \emph{categorical semantics}, which reveals the nature
of the computations -- rather than considering just the syntactic aspects of the type system.

\subsection{Properties and related notions}

The flat indexed comonad structure is all we need to give the semantics of the flat coeffect 
calculus. Before doing so in Section~\ref{sec:flat-semantics-calculus}, we briefly consider 
additional properties and other categorical structures that have been proposed mainly in the 
context of monads and effects and we look how they relate to indexed comonads.

\paragraph{Shape preservation.}
Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
this means that the core comonad structure does not provide a way of modeling computations where
the additional context changes during the computation. For example, in the \ident{NEList} comonad, 
the length of the list stays the same after applying \ident{cobind}.

Indexed comonads are not restricted by this property of comonads. For example, given the indexed
product comonad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$, the shape of
the context changes from providing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to providing
just implicit parameters $\cclrd{s}$.

@@

\subsection{When is coeffect not a monad}
Coeffect systems differ from effect systems in three important ways:

@@

This is, indeed, possible. Recall that the semantics of lambda abstraction in the flat
coeffect calculus is modelled using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. The operation takes
two contexts (wrapped in an indexed comonad $\ctyp{\cclrd{r}}\alpha$), combines their carried 
values and additional contextual information (implicit parameters). To obtain the (\emph{mabs}) 
rule, we can restrict the first parameter, which corresponds to the declaration site context:
%
\begin{equation*}
\begin{array}{rcll}
 \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{r}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{r \cup s}}(\alpha \times \beta)
 &\hspace{4em}(\textit{normal}) \\
 \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{\emptyset}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{s}}(\alpha \times \beta)
 &\hspace{4em}(\textit{restricted})
\end{array} 
\end{equation*}
%
In the (\emph{restricted}) version of the operation, the declaration site context requires
no implicit parameters and so all implicit parameters have to be satisfied by the call site.
The semantics using the restricted version corresponds to the (\emph{mabs}) rule shown above.

@@

@@

Next, we introduced the notion of \emph{flat indexed comonad}, which generalizes of comonads
and adds additional operations needed to provide categorical semantics of the flat coeffect 
calculus. The indices of the flat indexed comonad operations correspond to the
coeffect annotations in the type system and provide a foundation for the design of the 
calculus.

In the upcoming chapter, we move from \emph{flat} coeffect calculi, tracking whole-context
properties to \emph{structural} coeffect calculi, tracking per-variable information, thus
covering systems from the second half of Chapter~\ref{ch:applications}.

>>>>

\paragraph{Coeffects for liveness.}
As an example, we consider the third instance of coeffect calculus that was discussed in 
Chapter~\ref{ch:flat}. If we wanted to follow the development in the previous section for liveness, 
we would extend the target language with two kinds of expressions, \ident{Dead} representing a dead
context with no value and \ident{Live} representing a context with a value:
%
\begin{equation*}
\begin{array}{rcl}
e &=& \ldots \sep \ident{Dead} \sep \ident{Live}~e \\
\end{array}  
\end{equation*}
%
The typing rules promote the information about whether a value is available into the type-level 
and so a context carrying a live value is marked as $\ctyp{\cclrd{\ident{L}}}{\tau}$ while a
dead context has a type $\ctyp{\cclrd{\ident{D}}}{\tau}$.
%
\begin{equation*}
\tyrule{live}
  { \Gamma \vdash e:\tau }
  { \Gamma \vdash \ident{Live}~e : \ctyp{\cclrd{\ident{L}}}{\tau} }
\qquad
\tyrule{dead}
  { ~ }
  { \Gamma \vdash \ident{Dead} : \ctyp{\cclrd{\ident{D}}}{\tau} }
\end{equation*}
%
Finally, we need to add reduction rules that define the meaning of the comonadically-inspired
operations for liveness. Those follow the definitions given in Example~\ref{thm:semantics-indexed-opt}
when discussing the categorical semantics:
%
\begin{equation*}
\begin{array}{rl}
{\footnotesize(\emph{counit})} &
  \ident{counit}_{\cclrd{ \ident{L} }}~(\ident{Just}~v)   \rightarrow v
\\[0.5em]
{\footnotesize(\emph{cobind-1})} &
  \ident{cobind}_{\cclrd{\ident{L}}, \cclrd{\ident{L}}}~f~(\ident{Live}~v) \rightarrow \ident{Live}~(f~v)
\\
{\footnotesize(\emph{cobind-2})} &
  \ident{cobind}_{\cclrd{\ident{D}}, \cclrd{\ident{L}}}~f~(\ident{Live}~v) \rightarrow \ident{Live}~(f~\ident{Dead})
\\
{\footnotesize(\emph{cobind-3})} &
  \ident{cobind}_{\cclrd{\ident{L}}, \cclrd{\ident{D}}}~f~v \rightarrow \ident{Dead}
\\
{\footnotesize(\emph{cobind-4})} &
  \ident{cobind}_{\cclrd{\ident{D}}, \cclrd{\ident{D}}}~f~v \rightarrow \ident{Dead}
\end{array}
\end{equation*}
%
This language extension is safe because the reductions respect the typing of the
comonadically-inspired operations. The $\ident{counit}_\cunit$ reduction does not get stuck
for well-typed terms because $\cunit = \cclrd{\ident{L}}$ in the coeffect algebra and thus its 
argument is of type $\ctyp{\cclrd{\ident{L}}}{\tau}$ and will always be a value $\ident{Live}~v$.

Similarly, the when reducing $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$, the typing ensures that
the value passed as the second argument is of type $\ctyp{\cclrd{r}\cseq\cclrd{s}}{\tau_1}$.
In case of liveness, $\cclrd{r}\,\cseq\,\cclrd{s}=\cclrd{\ident{L}}$ if either
$\cclrd{r}=\cclrd{\ident{L}}$ or $\cclrd{s}=\cclrd{\ident{L}}$. This means that reductions
(\emph{cobind-1}) and (\emph{cobind-2}) will not get stuck because the value will be
$\ident{Live}~v$ and not \ident{Dead}. The reduction rules also preserve typing - the resulting
value is of type $\ctyp{\cclrd{s}}\tau_2$, that is $\ident{Live}~v$ for (\emph{cobind-1}),
(\emph{cobind-2}) and \ident{Dead} for (\emph{cobind-3}) and (\emph{cobind-4}).

\paragraph{Encoding liveness in Haskell.}
The liveness example can be encoded in Haskell using type-level features such as generalized
algebraic data types (GADTs) and type families \cite{haskell-faking,haskell-promotion,haskell-families}.
We do not aim to give a complete implementation, but to show that such encoding is possible and
would provide the necessary safety guarantees.

We first define types \ident{D} and \ident{L} to capture the coeffect annotations. 
Then we define a comonadically-inspired type $\ident{C}~r~a$ as a GADT with cases for
\ident{Live} and \ident{Dead} contexts. The type parameter $r$ represents a coeffect 
annotation:
%
\begin{equation*}
\begin{array}{l}
\kvd{data}~\ident{L} \\[-0.2em]
\kvd{data}~\ident{D} \\[0.5em]
\kvd{data}~\ident{C}~r~a~\kvd{where} \\[-0.2em]
\quad \ident{Live} \hspace{0.45em}:: a \rightarrow \ident{C}~\ident{L}~a\\[-0.2em]
\quad \ident{Dead} :: \ident{C}~\ident{D}~a
\end{array}
\end{equation*}
%
The definition matches with the typing rules (\emph{live}) and (\emph{dead}). The coeffect 
annotation for a live value is \ident{L} and the annotation for a dead value is \ident{D}.
To give the type of \ident{cobind}, we need a type-level function that encodes the operations
of the flat coeffect algebra. We model $\cseq$ as $\ident{Seq}~a~b$. The operation is defined
on types $\ident{L}$ and $\ident{D}$ and returns a type $\ident{D}$ if and only if both its
arguments are $\ident{D}$:
%
\begin{equation*}
\begin{array}{l}
\kvd{type}~\kvd{family}~\ident{Seq}~r~s :: *\\[-0.2em]
\quad \kvd{type}~\kvd{instance}~\ident{Seq}~\ident{D}~\ident{D} = \ident{D}\\[-0.2em]
\quad \kvd{type}~\kvd{instance}~\ident{Seq}~\ident{L}~s = \ident{L}\\[-0.2em]
\quad \kvd{type}~\kvd{instance}~\ident{Seq}~r~\ident{L} = \ident{L}
\end{array}
\end{equation*}
%
The \ident{counit} and \ident{cobind} operations can then be defined as Haskell functions
that have types corresponding to the typing rules (\emph{counit}) and (\emph{cobind}) given
in Figure~\ref{fig:semantics-ext}:
%
\begin{equation*}
\begin{array}{rcl}
\ident{counit} &::& \ident{C}~\ident{L}~a \rightarrow a \\[-0.2em]
\ident{cobind} &::& (\ident{C}~\ident{r}~a \rightarrow b) \rightarrow \ident{C}~(\ident{Seq}~r~s)~a \rightarrow \ident{C}~\ident{s}~b
\end{array}
\end{equation*}
%
Here, the additional type annotation ensures that \ident{counit} can only be called on a context
that contains a value and so calling the operation is not going to fail. Similarly, the type of the
\ident{cobind} operation now guarantees that if a function (used as the first argument) or the result
require a live context, it will be called with a value $\ident{C}~\ident{L}~a$ that is guaranteed
to contain a value.

\paragraph{Coeffects in dependently-typed languages.}
If we use the above encoding, type preservation is guaranteed by the type 
system of the target language. The equivalent of the progress property is guaranteed by the 
fact that the the implementation of the operations is well-defined.

@@

\section{Related categorical structures}
\label{sec:semantics-related}

Related work leading to coeffects has already been discussed in Chapter~\ref{ch:pathways} 
and we covered work related to individual concepts throughout the thesis. However, there is a
number of related categorical structures that are related to our \emph{indexed comonads}
(Section~\ref{sec:semantics-flat-idx}) that deserve additional discussion.

In Section~\ref{sec:semantics-related-related}, we discuss related approaches to adding 
indices to categorical structures (mostly monads). In Section~\ref{sec:semantics-related-monad},
we discuss a question that often arises when discussing coeffects and that is 
\emph{when is a coeffect (not) an effect?}

\subsection{Indexed categorical structures}
\label{sec:semantics-related-related}

Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
this means that the core comonad structure does not provide a way of modeling computations where
the additional context changes during the computation. For example, in the \ident{NEList} comonad, 
the length of the list stays the same after applying \ident{cobind}.

Indexed comonads are not restricted by this property of comonads. For example, given the indexed
product comonad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$, the shape of
the context changes from providing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to providing
just implicit parameters $\cclrd{s}$. 
Thus \emph{indexed comonads} are a generalization of \emph{comonads} that captures structures that
do not form a (non-indexed) comonad. However, indexing has been also discussed in the case of
\emph{monads.}

@@

\subsection{When is coeffect not a monad}
\label{sec:semantics-related-monad}

Coeffect systems differ from effect systems in three important ways:

@@

This is, indeed, possible. Recall that the semantics of lambda abstraction in the flat
coeffect calculus is modelled using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. The operation takes
two contexts (wrapped in an indexed comonad $\ctyp{\cclrd{r}}\alpha$), combines their carried 
values and additional contextual information (implicit parameters). To obtain the (\emph{mabs}) 
rule, we can restrict the first parameter, which corresponds to the declaration site context:

\begin{equation*}
\begin{array}{rcll}
 \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{r}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{r \cup s}}(\alpha \times \beta)
 &\hspace{4em}(\textit{normal}) \\
 \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{\emptyset}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{s}}(\alpha \times \beta)
 &\hspace{4em}(\textit{restricted})
\end{array} 
\end{equation*}
%
In the (\emph{restricted}) version of the operation, the declaration site context requires
no implicit parameters and so all implicit parameters have to be satisfied by the call site.
The semantics using the restricted version corresponds to the (\emph{mabs}) rule shown above.

@@

% ==================================================================================================
%
%     #####
%    #     # #    # #    # #    #   ##   #####  #   #
%    #       #    # ##  ## ##  ##  #  #  #    #  # #
%     #####  #    # # ## # # ## # #    # #    #   #
%          # #    # #    # #    # ###### #####    #
%    #     # #    # #    # #    # #    # #   #    #
%     #####   ####  #    # #    # #    # #    #   #
%
% ==================================================================================================

@@

In the previous chapter, we defined \emph{type system} for flat coeffect calculi that uniformly
captures the shared structure of context-aware computations. In this chapter, we completed
the unification by providing \emph{semantics} for flat coeffect calculi and proving the
\emph{safety} of coeffect languages for dataflow and implicit parameters. The semantics shown
here also guides the implementation that is discussed later in Chapter~\ref{ch:impl}.

The development presented in this chapter follows the well-known example of effects and monads.
We introduced the notion of \emph{indexed comonad}, which generalizes comonads
and adds additional operations needed to provide categorical semantics of the flat coeffect 
calculus and we demonstrated how implicit parameters, liveness and dataflow computations form
indexed comonads.

We then used the comonadic semantics to define a \emph{comonadically-inspired translation}
that turns programs written in a domain-specific coeffect language into a functional target 
language. This is akin to the Haskell do notation for monads. Finally, we extended the target
language with concrete implementation of comonadic operations for dataflow and implicit parameters
and we presented syntactic safety proof. In summary, the proof states that well-typed 
context-aware programs written in a coeffect language \emph{do not go wrong} (when translated
to a simple functional language and evaluated).

The proof relies on the fact that coeffect annotations (provided by the coeffect type system)
guarantee that the required context is available in the comonadic value that represents the 
context and we also discussed how this would guarantee safety in languages with sufficiently
expressive type system such as Haskell.

In the following chapter, we move from \emph{flat} coeffect calculi, tracking whole-context
properties to \emph{structural} coeffect calculi, tracking per-variable information, thus
covering systems from the second half of Chapter~\ref{ch:applications}.
